---
title: "Cylistic Project"
author: "Pritam Jal"
date: "19/11/2021"
output: html_document
---

# STEP:1 - Installing the necessary Packages


"tidyverse" package will help wrangle data.
"lubridate" package will help helps wrangle date attributes.
"skimr" package will help Us get summary data.
"janitor" package is used for examining and cleaning dirty data.  
"dplyr" package is used for data manipulation.

```{r}
```

# STEP:2 - Loading the Packages
```{r}
library(tidyverse) 
library(lubridate) 
library(skimr) 
library(janitor)
library(dplyr)

```

# STEP:3 - COLLECT THE DATA 

```{r}
setwd("C:/Users/PC-032/Desktop/Cyclistic 12 months data/CSV")
Trips_Sep20<- read_csv ('202009-divvy-tripdata.csv')
Trips_Oct20<- read_csv ('202010-divvy-tripdata.csv')
Trips_Nov20<- read_csv ('202011-divvy-tripdata.csv')
Trips_Dec20<- read_csv ('202012-divvy-tripdata.csv')
Trips_Jan21<- read_csv ('202101-divvy-tripdata.csv')
Trips_Feb21<- read_csv ('202102-divvy-tripdata.csv')
Trips_Mar21<- read_csv ('202103-divvy-tripdata.csv')
Trips_Apr21<- read_csv ('202104-divvy-tripdata.csv')
Trips_May21<- read_csv ('202105-divvy-tripdata.csv')
Trips_June21<- read_csv ('202106-divvy-tripdata.csv')
Trips_July21<- read_csv ('202107-divvy-tripdata.csv')
Trips_Aug21<- read_csv ('202108-divvy-tripdata.csv')

```

# STEP:4 -  WRANGLE DATA AND COMBINE INTO A SINGLE FILE 

## Compare column names each of the files, the column name should match perfectly so that we combine all the files

```{r}
colnames(Trips_Sep20 )
colnames(Trips_Oct20 )
colnames(Trips_Nov20 )
colnames(Trips_Dec20 )
colnames(Trips_Jan21 )
colnames(Trips_Feb21 )
colnames(Trips_Mar21 )
colnames(Trips_Apr21 )
colnames(Trips_May21 )
colnames(Trips_June21 )
colnames(Trips_July21 )
colnames(Trips_Aug21 )
```
## Inspect the dataframes and look for inconguencies
```{r}
str(Trips_Sep20 )
str(Trips_Oct20 )
str(Trips_Nov20 )
str(Trips_Dec20 )
str(Trips_Jan21 )
str(Trips_Feb21 )
str(Trips_Mar21 )
str(Trips_Apr21 )
str(Trips_May21 )
str(Trips_June21 )
str(Trips_July21 )
str(Trips_Aug21 )
```


## We need to compare column datatype of all dataframe by using compare_df_cols, so that there will be no issue in joining files. 

```{r}
compare_df_cols (Trips_Sep20, Trips_Oct20, Trips_Nov20, Trips_Dec20, Trips_Jan21, Trips_Feb21, Trips_Mar21, Trips_Apr21, Trips_May21, Trips_June21, Trips_July21, Trips_Aug21,  return = "mismatch")
```


## Convert end_station_id and start_station_id into character so that they can be stack correctly 

```{r}
Trips_Sep20 <- mutate(Trips_Sep20, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_Oct20 <- mutate(Trips_Oct20, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_Nov20 <- mutate(Trips_Nov20, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_Dec20 <- mutate(Trips_Dec20, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_Jan21 <- mutate(Trips_Jan21, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_Feb21 <- mutate(Trips_Feb21, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_Mar21 <- mutate(Trips_Mar21, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_Apr21 <- mutate(Trips_Apr21, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_May21 <- mutate(Trips_May21, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_June21 <- mutate(Trips_June21, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_July21 <- mutate(Trips_July21, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_Aug21 <- mutate(Trips_Aug21, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
```



## Recheck if the Column names to see if all matches.

```{r}
compare_df_cols (Trips_Sep20, Trips_Oct20, Trips_Nov20, Trips_Dec20, Trips_Jan21, Trips_Feb21, Trips_Mar21, Trips_Apr21, Trips_May21, Trips_June21, Trips_July21, Trips_Aug21,  return = "mismatch")
```

## Stack individual dataframes into one big dataframe 

```{r}
all_trips <- bind_rows(Trips_Sep20, Trips_Oct20, Trips_Nov20, Trips_Dec20, Trips_Jan21, Trips_Feb21, Trips_Mar21, Trips_Apr21, Trips_May21, Trips_June21, Trips_July21, Trips_Aug21)
```

## Remove unused column 
```{r}
all_trips <- all_trips %>% 
 select(-c(start_lat, start_lng, end_lat, end_lng))
```

## Rename Columns 

```{r}
all_trips <- all_trips %>% 
  rename( trip_id= ride_id,
          ride_type =rideable_type, 
          start_time = started_at,
          end_time =ended_at,
          from_station_name = start_station_name,
          from_station_id = start_station_id,
          to_station_name = end_station_name,
          to_station_id = end_station_id,
          usertype = member_casual)

```

# STEP:5 - CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS 

##  Inspect the new table that has been created 
```{r}
colnames(all_trips) #List of column names
```

```{r}
dim(all_trips) #Dimensions of the data frame?
```

```{r}
str(all_trips) #See list of columns and data types (numeric, character, etc)
```

```{r}
summary(all_trips) #Statistical summary of data. Mainly for numerics
```



## This will allow us to aggregate ride data for each month, day, or year before completing these operations we could only aggregate at the ride level
```{r}
all_trips$date <- as.Date(all_trips$start_time) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```


## Add a “ride_length” calculation to all_trips (in seconds) 
```{r}
all_trips$ride_length <- difftime(all_trips$end_time,all_trips$start_time)

```


##  Convert “ride_length” from Factor to numeric so we can run calculations on the data 
```{r}
is.factor(all_trips$ride_length)
```

```{r}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```


## Remove “bad” data The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative

```{r}
skim(all_trips$ride_length)
```


```{r}
all_trips_v2 <- all_trips[!(all_trips$ride_length<0),]
skim(all_trips_v2)
```


# STEP:6 -  CONDUCT DESCRIPTIVE ANALYSIS 


## Descriptive analysis on ride_length (all figures in seconds) 
```{r}
summary(all_trips_v2$ride_length)
```



# STEP:7 -  Export to CSV file for further analysis 
```{r}
write.csv(all_trips_v2, "data.csv")
```


